---
title: "P01. Introduction to R, part 1"
---

A. Create simple objects of different types in the workspace

```{r eval = F}
item1 <- 1
item2 <- "a"
item3 <- 3.78901
```

B. Objects with more than 1 entry are called vectors

```{r eval = F}
# create some vectors using c() (short for concatenate)
# all items of a vector must be the same class

object1 <- c("a", "b", "c")
object2 <- 1:3
object3 <- c(1.3, -4.5, 6.99)

# you can create vectors using other named items or objects
object4 <- c(item1, item2, item3)

# take a look
object1
object2
object3
object4

```

C. Look for your objects in the environment tab (upper right)

D. What class of objects is each object?

Answer:

```{r eval = F}
# use class() to find this
class(object1)
class(object2)
class(object3)
class(object4)
```

Is object 4 the class you expected?

Answer:

E. You can manipulate objects and make operations on them

```{r eval = F}
object2 + 1
object3*object3
object2/object3
object1 + object2
```

what happened at each of these commands? is it what you expected?

Answer:  
F. You can view the history of commands that were run.
check the history of commands in the history tab (upper right)

G. Make some other objects

```{r eval = F}
df1 <- data.frame(ID=1:5,
animal=c("bear", "cat", "horse", "cat", "pig"),
weight=c(200, 5, 600, 8, 100))
mat1 <- matrix(data=1:50, nrow=10, ncol=5)

# view these objects
df1
mat1
View(df1)
View(mat1)

```

In R, what is the difference between a data.frame and a matrix? hint: google it.

Answer:

H. Calculate the mean weight of the animals in df1. (i.e. mean of column 3 of df1)

```{r eval = F}
# you can either refer to a column by name or by index
# check what the names are
colnames(df1)

# reference a column by name
df1$weight

# check how to use the function "mean"
# ? allows you to view the help file of any inbuilt function
?mean

# calculate the mean
mean(df1$weight)
mean(df1[, 3])
```

Answer:

I. does mat1 have column names?

```{r eval = F}
# set column names for mat1
colnames(mat1) <- c("A", "B", "C", "D", "E")
mat1


```
Answer:

J. What is the value of the element row 5, column C in mat1? what is the value of row 8, column E?

```{r eval = F}
# inspect the element by referencing the row, then column, either by name or number
mat1[5, "C"]
mat1[5, 3]

```
Answer:

K. Discard all of mat1 except the 1st and 4th column.
```{r eval = F}
# subset mat1 to keep only the 1st and 4th column by number or name
mat1[ , c(1, 4)]
mat1[ , c("A", "D")]

# Create a new object of the smaller matrix.
# assign the subsetted version as a new object, mat2
mat2 <- mat1[ , c("A", "D")]
mat2 <- mat1[ , c(1, 4)]
```