---
title: "04. Ordinary differential equations (ODEs): Solutions"
---

# ODE's in R: Practical

**Click [here](04_ODEs_practical.qmd) to return to the practical.**

## Solving the SI model using `deSolve` 

The code below will solve the SI model with beta = 0.4 over 50 days with initial conditions S(0) = 99, I(0) = 1. Run the following lines including the plot() line to create a plot of the output

```{r}
library(deSolve) # Load in the deSolve package. If the package is not installed, install using the install.packages() function
# Define model function

SI_model <- function(times, state, parms){
  ## Define variables
  S <- state["S"]
  I <- state["I"]
  N <- S + I
  # Extract parameters
  beta <- parms["beta"]
  # Define differential equations
  dS <- - (beta * S * I) / N
  dI <- (beta * S * I) / N
  res <- list(c(dS, dI))
  return(res)
}

# Define parameter values
parameters <- c(beta = 0.4)

# Define time to solve equations
times <- seq(from = 0, to = 50, by = 1)

# Define initial conditions
N <- 100
I_0 <- 1
S_0 <- N - I_0
state <- c( S = S_0, I = I_0)

# Solve equations
output_raw <- ode(y = state, times = times, func = SI_model, parms = parameters,
                  method = "rk4")
# Convert to data frame for easy extraction of columns
output <- as.data.frame(output_raw)

# Plot output
par( mfrow = c(1, 1))
plot( output$time, output$S, type = "l", col = "blue", lwd = 2, ylim = c(0, N),
      xlab = "Time", ylab = "Number")
lines( output$time, output$I, lwd = 2, col = "red", type = "l")
legend("topright", legend = c("Susceptible", "Infected"),
       lty = c(1, 1), col = c("blue", "red"), lwd = 2, bty = "n")
```

### Question 1:

#### 1a) Increase the initial number of infected individuals, what happens to the output?

**Answer:** The number of infecteds as a higher intercept but same asymptote

#### 1b) What does the 'by' argument in the 'times' vector represent? 

Answer: The time steps of the model solution

#### 1c) Increase the value of the "by" argument. What happens to the output? HINT: plot using type = "b" to plot both lines and points

Answer: Solution becomes less precise.

## Solving the SIR model using `deSolve`

Now, let's solve a Susceptible-Infected-Recovered set of ordinary equations. There are now states; Susceptible, Infected, Recovered. Once infected people recover at a rate gamma = 0.1. The code for the model function ONLY is given below.

```{r}
# Define model function 
SIR_model <- function(times, state, parms){
  ## Define variables
  S <- state["S"]
  I <- state["I"]
  R <- state["R"]
  N <- S + I + R
  # Extract parameters
  beta <- parms["beta"]
  gamma <- parms["gamma"]
  # Define differential equations
  dS <- - (beta * S * I) / N
  dI <- (beta * S * I) / N - gamma * I
  dR <- gamma * I
  res <- list(c(dS, dI, dR ))
  return(res)
}
```

### Question 2:

#### 2a) To solve this model, you need to write a new vector of parameters and of state variables. You can assume that there are no recovered individuals at time 0 and 1 infected. Plot the output of the SIR model with different colours for Susceptible, Infected and Recovered individuals.

```{r}
# Define parameter values
parameters <- c(beta = 0.4, gamma = 0.1)

# Define initial conditions
N <- 100
I_0 <- 1
S_0 <- N - I_0
R_0 <- 0
state <- c(S = S_0, I = I_0, R = R_0)

# Solve equations
output_raw <- ode(y = state, times = times, func = SIR_model, parms = parameters,
                  method = "rk4")
# Convert to data frame for easy extraction of columns
output <- as.data.frame(output_raw)

# Plot output
par(mfrow = c(1, 1))
plot(output$time, output$S, type = "l", col = "blue", lwd = 2, ylim = c(0, N),
     xlab = "Time", ylab = "Number")
lines(output$time, output$I, lwd = 2, col = "red")
lines(output$time, output$R, lwd = 2, col = "green")
legend( "topright", legend = c("Susceptible", "Infected", "Recovered"),
        lty = 1, col = c("blue", "red", "green"), lwd = 2, bty = "n")

```

#### 2b) Change the value of the transmission rate so that the basic reproduction number is less than one, i.e. R_0 \< 1, what happens to the output? 

Hint: Recall that for an SIR model, the basic reproduction number, R_0 = beta / gamma

**Answer:** The epidemic does not take off

```{r}
parameters <- c(beta = 0.09, gamma = 0.1)

output_raw <- ode(y = state, times = times, func = SIR_model, parms = parameters,
                  method = "rk4")
# Convert to data frame for easy extraction of columns
output <- as.data.frame(output_raw)

# plot results
par(mfrow = c(1, 1))
plot( output$time, output$S, type = "l", col = "blue", lwd = 2, ylim = c(0, N),
      xlab = "Time", ylab = "Number")
lines(output$time, output$I, lwd = 2, col = "red")
lines(output$time, output$R, lwd = 2, col = "green")
legend("topright", legend = c("Susceptible", "Infected", "Recovered"),
       lty = 1, col = c("blue", "red", "green"), lwd = 2, bty = "n")

```

## Solving the SEIR model using `deSolve`

The code below has been written to solve a Susceptible-Exposed-Infected-Recovered model. In the model individuals can be in one of four states. Once infected, susceptible individuals move to the exposed class. Exposed individuals become infectious at a rate delta = 0.14.

### Question 3:

#### 3) If you attempt to run the code below, you will receive an error when `ode()` is called. Fix the SEIR_model function so that the ode() line runs without errors.

```{r}
# Define model function 
SEIR_model <- function(times, state, parms){
  ## Define variables
  S <- state[1]
  E <- state[2]
  I <- state[3]
  R <- state[4]
  N <- S + E + I + R
  # Extract parameters
  beta <- parms["beta"]
  gamma <- parms["gamma"]
  delta <- parms["delta"]
  # Define differential equations
  dS <- - (beta * S * I ) / N
  dE <- (beta * S * I) / N - delta * E
  dI <- delta * E - gamma * I
  dR <- gamma * I
  res <- list(c(dS, dE, dI, dR))
  return(res)
}

# Define parameters  
parameters <- c( beta = 0.4, gamma = 0.1, delta = 0.14)

# Define time to run model
times <- seq(from = 0, to = 50, by = 1)

# Define initial conditions
N <- 100
I_0 <- 1
S_0  <- N - I_0
R_0 <- 0
state <- c(S = S_0, E = 0, I = I_0, R = R_0)

# Solve equations
output_raw <- ode(y = state, times = times, func = SEIR_model, parms = parameters,
                  method = "rk4")
# Convert to data frame for easy extraction of columns
output <- as.data.frame(output_raw)

# plot results
par(mfrow = c(1, 1))
plot( output$time, output$S, type = "l", col = "blue", lwd = 2, ylim = c(0, N),
      xlab = "Time", ylab = "Number")
lines(output$time, output$I, lwd = 2, col = "red")
lines(output$time, output$R, lwd = 2, col = "green")
lines(output$time, output$E, lwd = 2, col = "cyan")
legend("topright", legend =  c("Susceptible", "Exposed", "Infected", "Recovered"),
       lty = 1, col = c("blue", "red", "green", "cyan"), lwd = 2, bty = "n")

```

### Optional questions: adding vaccination to the SIR model

#### 4) Extend the SIR model to include a vaccinated class. 

Here we assume that susceptible individuals are vaccinated at a rate v = 0.05. The vaccine is 100% effective, so once vaccinated, individuals cannot become infected. HINT: you will need to create a new class V, you can assume that the initial number of vaccinated individuals is 0.

```{r}
# Define model function 
SIRV_model <- function(times, state, parms){
  ## Define variables
  S <- state["S"]
  I <- state["I"]
  R <- state["R"]
  V <- state["V"]
  N <- S + I + R + V
  # Extract parameters
  beta <- parms["beta"]
  gamma <- parms["gamma"]
  v <- parms["v"]
  # Define differential equations
  dS <- - (beta * S * I) / N - v * S
  dI <- (beta * S * I) / N - gamma * I
  dR <- gamma * I
  dV <- v * S
  res <- list(c(dS, dI, dR, dV))
  return(res)
}


# Define parameter values
parameters <- c(beta = 0.4, gamma = 0.1, v = 0.05)

# Define initial conditions
N <- 100
I_0 <- 1
S_0 <- N - I_0
R_0 <- 0
V_0 <- 0
state <- c(S = S_0, I = I_0, R = R_0, V = V_0)

# Solve equations
output_raw <- ode(y = state, times = times, func = SIRV_model, parms = parameters,
                  method = "rk4")
# Convert to data frame for easy extraction of columns
output <- as.data.frame(output_raw)

# Plot results
par(mfrow = c(1, 1))
plot(output$time, output$S, type = "l", col = "blue", lwd = 2, ylim = c(0, N),
     xlab = "Time", ylab = "Number")
lines(output$time, output$I, lwd = 2, col = "green")
lines(output$time, output$R, lwd = 2, col = "red")
lines(output$time, output$V, lwd = 2, col = "purple")
legend( "topright", legend = c("Susceptible", "Infected", "Recovered", "Vaccinated"),
        lty=1, col = c("blue", "red", "green", 'purple'), lwd = 2, bty = "n")
```
